Density dependence

### 2020-10-05 R/density dependence.Rmd
prototype of excluding predicted values too far from data

```{r}
ha <-
  ha %>%
  ungroup() %>% 
  arrange(ha_id_number, year) %>% 
  #scale year, and convert to factor since I'll be using as random effect
  mutate(year_post = as_factor(year - min(year)))
```

Use just 1-ha and cf

```{r}
ha_sub <-
  ha %>% 
  filter(habitat %in% c("CF", "1-ha")) %>% 
  mutate(log_shts = log(shts),
         log_shts_next = log(shts_next),
         habitat = as.factor(habitat),
         ha_id_number = as.factor(ha_id_number))
```

@@

```{r}
mpois <-
  glmer(
    shts_next ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
    family = poisson(link = "log"),
    data = ha_sub
  )

@@

mnorm <-
  glmer(
    shts_next ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
    family = gaussian(link = "log"),
    data = ha_sub
  )
```

@@

```{r}
m_growth2 <- update(m_growth, .~. - (1|habitat:plot))
AICtab(m_growth, m_growth2)
```

@@

No density dependence

@@

```{r}
m_surv <-
  glmer(
    surv ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) + (1|year_post) + (1|habitat:plot), 
    family = binomial, data = ha_sub
  )
```

@@

```{r}
m_surv2 <- update(m_surv, .~. - (1|habitat:plot))
AICtab(m_surv, m_surv2)
```

@@

Positive density dependence for survival.

@@

```{r}
g_growth <-
  gam(shts_next ~ 
        s(log_shts) + #Essentially linear, could replace with parametric
        habitat +
        s(Q, L,
          bs = "cb",
          xt = list(bs = "cr")), #+
        # s(ha_id_number, bs = "re"), #this takes waaay too long to converge
      family = gaussian(link = "log"),
      data = ha_sub,
      method = "REML")
```

@@

```{r}
pred_growth <- pred_cb(Q, L, g_growth)
ggplot(pred_growth, aes(x = x, y = lag, z = fitted, fill = fitted)) +
  geom_raster(interpolate = TRUE) +
  geom_contour(color = "grey50", size = 0.4) + 
  scale_y_continuous("distance (m)", expand = c(0,0)) +
  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A")
```

@@

```{r}
g_surv <-
  gam(surv ~
        s(log_shts) +
        habitat +
        s(Q, L,
          bs = "cb",
          # k = c(10, 10),
          xt = list(bs = "cr")),
      family = binomial,
      data = ha_sub,
      method = "REML")
```

@@

```{r}
pred_surv <- pred_cb(Q, L, g_surv)
ggplot(pred_surv, aes(x = x, y = lag, z = fitted, fill = fitted)) +
  geom_raster(interpolate = TRUE) +
  geom_contour(color = "grey50", size = 0.4) + 
  scale_y_continuous("distance (m)", expand = c(0,0)) +
  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A")
```

>>>>

```{r}
ha <-
  ha %>%
  ungroup() %>% 
  arrange(ha_id_number, year) %>% 
  #scale year, and convert to factor since I'll be using as random effect
  mutate(year_post = as_factor(year - min(year)),
         log_shts = log(shts),
         log_shts_next = log(shts_next),
         habitat = as.factor(habitat),
         ha_id_number = as.factor(ha_id_number),
         plot = as.factor(plot))
```

Emilio suggested starting with a couple of small subsets of data to make sure I'm getting things right.

1. Just continuous forest
2. Just a single plot (e.g. 5750)
3. Just plants with the median # of shoots (e.g. 3)

```{r}
# ha_sub <-
#   ha %>%
#   filter(habitat %in% c("CF", "1-ha"))

# ha_sub <-
#   ha %>%
#   filter(habitat == "CF")

# ha_sub <-
#   ha %>%
#   filter(plot == 5750)

# ha_sub <-
#   ha %>%
#   filter(shts == median(shts, na.rm = TRUE))

ha_sub <- ha
```

```{r}
ha_sub %>% 
  group_by(year, plot) %>% 
  summarize(n_distinct(row), n_distinct(column))
```
There's something funky going on in 2009 with row and column positions. I opened an issue and this will get fixed eventually.

@@

```{r}
mpois <-
  glmer(
    shts_next ~
      log_shts + 
      habitat +
      shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) +
      (1|plot) +
      (1|year_post),
    family = poisson(link = "log"),
    data = ha_sub
  )

@@

mnorm <-
  glmer(
    shts_next ~ 
      log_shts +
      habitat +
      shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) +
      (1|plot) +
      (1|year_post),
    family = gaussian(link = "log"),
    data = ha_sub
  )
```

@@

```{r}
m_growth2 <- update(m_growth, .~. - (1|plot))
AICtab(m_growth, m_growth2)
```

@@

- Just 5750: marginaly significant negative density dependence at 2.5--5m
- Only CF: no significant effect of density dependence
- Only 3-shoot plants: no significant effect of density dependence
- Full dataset: No density dependence

@@

```{r}
m_surv <-
  glmer(
    surv ~
      log_shts +
      habitat +
      shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) +
      (1|plot) +
      (1|year_post), 
    family = binomial, data = ha_sub
  )
```

@@

```{r}
m_surv2 <- update(m_surv, .~. - (1|plot))
AICtab(m_surv, m_surv2)
```

@@

- Just 5750: no density dependence
- Only CF: no density dependence
- Only 3-shoot plants: Marginally significant density dependence.  Negative at 0--2.5m, positive at 2.5--5m
- Full dataset: no density dependence

@@

```{r}
g_growth <-
  gam(shts_next ~ 
        # s(ha_id_number, bs = "re") + #this takes waaay too long to converge
        s(plot, bs = "re") +
        log_shts + #when modeled as a spline, edf is essentially a line
        habitat +
        s(Q, L,
          bs = "cb",
          xt = list(bs = "cr")),
      family = gaussian(link = "log"),
      data = ha_sub,
      method = "REML")
```

@@

```{r}
pred_growth <- pred_cb(Q, L, g_growth)
```

Which predicted values are too far from data to sensibly plot?  (this is modeled after `mgcv::exclude.too.far`, which does not work for these models)

```{r}
dat <-
  ha_sub %>%
  select(shts_0.5_1:shts_4.5_5) %>% 
  pivot_longer(everything(),
               names_prefix = "shts_",
               names_sep = "_",
               names_to = c("inner", "outer"),
               names_transform = list(inner = as.numeric, outer = as.numeric),
               values_to = "x") %>% 
  mutate(lag = (inner + outer) / 2)

which.too.far <- function(grid, data, x, y, dist = 0.1) { 
  x <- enquo(x)
  y <- enquo(y)
  # x <- quo(x)
  # y <- quo(lag)

  grid1 <-
    grid %>% 
    mutate(min_g_x = min(!!x, na.rm = TRUE),
           min_g_y = min(!!y, na.rm = TRUE),
           g_x = !!x - min_g_x,
           g_y = !!y - min_g_y) %>% 
    mutate(max_g_x = max(g_x, na.rm = TRUE),
           max_g_y = max(g_y, na.rm = TRUE),
           g_x = g_x / max_g_x,
           g_y = g_y / max_g_y)

  data1 <-
    data %>% 
    mutate(d_x = (!!x - first(grid1$min_g_x)) / first(grid1$max_g_x),
           d_y = (!!y - first(grid1$min_g_y)) / first(grid1$max_g_y))

  if (dist < 0)
    stop("supplied dist negative")

  #where dat is a 2-column matrix of x and y coords of the true data used to build the model
  min_dist <- function(g_x, g_y, dat) {
    dat[,1] <- dat[,1] - g_x
    dat[,2] <- dat[,2] - g_y
    min(
      sqrt(
        (dat[,1]^2 + dat[,2]^2)
      ), na.rm = TRUE
    )
  }
  o <- 
    grid1 %>%
    rowwise() %>%
    mutate(min_dist = min_dist(g_x, g_y, cbind(data1$d_x, data1$d_y)),
           too.far = min_dist > dist) %>% 
    select(-g_x, -g_y, -max_g_x, -max_g_y, -min_g_x, -min_g_y)
  return(o)
}

pred_growth <- which.too.far(pred_growth, dat, x = x, y = lag)
```

```{r}
ggplot(pred_growth %>% filter(min_dist < 0.1),
       aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  geom_contour(color = "grey50", size = 0.4) + 
  scale_y_continuous("distance (m)", expand = c(0,0)) +
  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A")
```

Plot a "slice" through 5m

```{r}
ggplot(pred_growth %>% filter(lag == 0.75),
       aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
  geom_line() + 
  geom_ribbon(alpha = 0.3) +
  geom_point(data = ha_sub, aes(x = shts_0.5_1, y = shts_next), inherit.aes = FALSE) +
  coord_cartesian(ylim = c(0,20))
```

- TODO: Try with Florestal (really huge plants there)
- 5750 only: essentially flat surface, except for positive density dependence at 5m.  Seems spurious, would want to see CI.
- CF only: similar.  Essentially flat except for impossibly large positive density dependence at 5 m (expected value of 200+ shoots at highest conspecific density)
- 3-shoot plants only: No significant effect of density.
- Full dataset: Significant, but same pattern.  Large effects only at 5m.  Probably driven by a few extreme values of high-density plots.

@@

```{r}
g_surv <-
  gam(surv ~
        s(plot, bs = "re") +
        s(log_shts) +
        habitat +
        s(Q, L,
          bs = "cb",
          xt = list(bs = "cr")),
      family = binomial,
      data = ha_sub,
      method = "REML")
```

@@

```{r}
pred_surv <- pred_cb(Q, L, g_surv)
ggplot(pred_surv, aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  geom_contour(color = "grey50", size = 0.4) + 
  scale_y_continuous("distance (m)", expand = c(0,0)) +
  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A")
```

- 5750 only: No significant density dependence
- CF only: Also not significant.
- 3-shoot plants only: no significant density dependence
- Full dataset: no significant effect of density dependence.
